Typing a linear \picalc

### 2020-03-31 paper/paper.tex
Draft on leftover typing, typesetting improvements

\newcommand{\type}[1]{\textcolor{blue}{\mathtt{\operatorname{#1}}}}
\newcommand{\constr}[1]{\textcolor{olive}{\mathtt{\operatorname{#1}}}}
\newcommand{\field}[1]{\textcolor{red}{\mathtt{\operatorname{#1}}}}

@@

% Functions
\newcommand{\subst}[3]{#1[#2/#3]}

@@

% Types
\newcommand{\Set}{\type{SET}}
\newcommand{\reduce}[1]{\; \type{\longrightarrow}_{#1} \;}
\newcommand{\types}[4]{#1 \; \type{\propto} \; #2 \; \type{\vdash} \; #3 \; \type{\boxtimes} \; #4}
\newcommand{\contains}[6]{#1 \; \type{\propto} \; #2 \; \type{_[} _{#3} \type{_]=} \; #4 \; \type{\propto} \; #5 \; \type{\boxtimes} \; #6}
\newcommand{\Var}{\type{VAR}}
\newcommand{\Process}{\type{PROCESS}}
\newcommand{\Unused}{\type{UNUSED}}
\newcommand{\PreCtx}{\type{PRECTX}}
\newcommand{\Ctx}{\type{CTX}}
\newcommand{\Type}{\type{TYPE}\;}
\newcommand{\Idx}{\type{IDX}\;}
\newcommand{\Idxs}{\type{IDXS}}
\newcommand{\Carrier}{\type{CARRIER}}
\newcommand{\N}{\type{\mathbb{N}}}
\newcommand{\Channel}{\type{CHANNEL}}
\newcommand{\Rec}{\type{REC}}
\newcommand{\Quantifier}{\type{QUANTIFIER}}
\newcommand{\eq}[1]{\; \type{=}_{#1} \;}
\newcommand{\eqeq}{\; \type{\equiv} \;}

@@

\begin{nidefinition}
  An indexed set of partial commutative monoids is a set of indices such that for every index there exists a carrier $C$ with a $\Quantifier_C$ algebra defined on it.
  This allows for several algebras to coexist in the same type system.
  Figure \ref{fig:indexed-multiplicities} captures these requirements: each index $i$ has a corresponding carrier type $\field{CARRIER}_i$ such that $\field{CARRIER}_i$ satisfies the algebra $\Quantifier_{\field{CARRIER}_i}$.
  We require the index type to be inhabited -- necessary to prove type preservation on $\new \PO_{\suc n} \eqeq \PO_n$.
  \todo{lift op to $^2$}

@@

@@

  \todo{example}
  \todo{lift op to $^2$}

@@

    \inferrule
        {t : \Type \\
          x \; y \; z : \Carrier_{idx}^2\\
          \op{x}{y}{z}}
    {\constr{zero} : \contains{\gamma \comma t}{\Gamma \comma x}{zero}{t}{y}{\Gamma \comma z}}

@@

\lemma{The typing context $\gamma$ at index $i$ is type $t$.}
\lemma{The input multiplicity context $\Gamma$ at index $i$ is the result of performing the monoidal operation on $x$ and the output multiplicity context $\Delta$ at index $i$.}
\lemma{For every index $j$ other than $i$, the input multiplicity context $\Gamma$ at index $i$ is the output multiplicity $\Delta$ at index $i$.}
\corollary{Given the witness $\contains{\gamma}{\Gamma}{i}{t}{x}{\Delta}$, $\Gamma \equiv \Delta$ iff $x \equiv \zero$.}
\proof{By structural induction on the variable reference}

\subsection{Typing judgments}\label{typing-judgments}

The indices in the variable reference rules are then used as the variable references of the underlying process.

\todo{comment on balancing in chan rule}

\begin{nidefinition}
  \begin{figure}[h]
  \begin{mathpar}
    \datatype{
      \gamma : \PreCtx_n \\
      \stacked{
        idxs : \Idxs_n \\\\
        \Gamma : \Ctx_{idxs}} \\
      P : \Process_n \\
      \Delta : \Ctx_{idxs}}
    {\types{\gamma}{\Gamma}{P}{\Delta} : \Set}
    \; \textsc{Types}

@@

    \inferrule
    {l : \types{\gamma}{\Gamma}{P}{\Delta} \\\\
     r : \types{\gamma}{\Delta}{Q}{\Xi}}
    {\constr{comp} \; l \; r : \types{\gamma}{\Gamma}{\comp{P}{Q}}{\Xi}}
  \end{mathpar}
  \caption{This is...}
  \end{figure}
\end{nidefinition}

@@

@@

\item [Affine types]

@@

>>>>

\newcommand{\type}[1]{\textcolor{blue}{\operatorname{#1}}}
\newcommand{\constr}[1]{\textcolor{olive}{\operatorname{#1}}}
\newcommand{\field}[1]{\textcolor{red}{\operatorname{#1}}}

@@

% Functions
\newcommand{\subst}[3]{#1[#2/#3]}
\newcommand{\opsquared}[3]{#1 \, \coloneqq \, #2 \, \cdot^2 \, #3}
\newcommand{\opctx}[3]{#1 \, \coloneqq \, #2 \, \otimes \, #3}

@@

% Types
\newcommand{\Set}{\type{SET}}
\newcommand{\reduce}[1]{\; \type{\longrightarrow}_{#1} \;}
\newcommand{\types}[4]{#1 \; \type{\propto} \; #2 \; \type{\vdash} \; #3 \; \type{\boxtimes} \; #4}
\newcommand{\contains}[6]{#1 \; \type{\propto} \; #2 \; \type{\ni}_{#3} \; #4 \; \type{\propto} \; #5 \; \type{\boxtimes} \; #6}
\newcommand{\Var}{\type{VAR}}
\newcommand{\Process}{\type{PROCESS}}
\newcommand{\Unused}{\type{UNUSED}}
\newcommand{\PreCtx}{\type{PRECTX}}
\newcommand{\Ctx}{\type{CTX}}
\newcommand{\Type}{\type{TYPE}\;}
\newcommand{\Idx}{\type{IDX}\;}
\newcommand{\Idxs}{\type{IDXS}}
\newcommand{\Carrier}{\type{CARRIER}}
\newcommand{\N}{\type{\mathbb{N}}}
\newcommand{\Channel}{\type{CHANNEL}}
\newcommand{\Rec}{\type{REC}}
\newcommand{\Quantifier}{\type{QUANTIFIER}}
\newcommand{\eq}[1]{\; \type{=}_{#1} \;}
\newcommand{\eqeq}{\; \type{\equiv} \;}

@@

\begin{note}
  We will often work with input and output multiplicities pairs, and thus we use the notation $\Carrier^2$ to stand for the type $\Carrier \times \Carrier$, and $\opsquared{x}{y}{z}$ to stand for a monoidal operation on pairs of multiplicities -- where all of its properties are trivially lifted.
\end{note}

\begin{nidefinition}
  An indexed set of partial commutative monoids is a set of indices such that for every index there exists a carrier $C$ with a $\Quantifier_C$ algebra defined on it.
  This allows for several algebras to coexist in the same type system.
  Figure \ref{fig:indexed-multiplicities} captures these requirements: each index $i$ has a corresponding carrier type $\field{CARRIER}_i$ such that $\field{CARRIER}_i$ satisfies the algebra $\Quantifier_{\field{CARRIER}_i}$.
  We require the index type to be inhabited -- necessary to prove type preservation on $\new \PO_{\suc n} \eqeq \PO_n$.

@@

\begin{note}
  We lift the monoidal operation on pairs of multiplicities $\opsquared{x}{y}{z}$ to a monoidal operation on contexts with the same underlying vector of indices $\opctx{\Gamma}{\Delta}{\Xi}$, where all the properties of the partial monoid are trivially lifted to contexts.
\end{note}

@@

  \todo{example}

@@

    \inferrule
        {t : \Type \\
          x \; y \; z : \Carrier_{idx}^2\\
          \opsquared{x}{y}{z}}
    {\constr{zero} : \contains{\gamma \comma t}{\Gamma \comma x}{zero}{t}{y}{\Gamma \comma z}}

@@

\begin{lemma}
  The typing context $\gamma$ at index $i$ is type $t$.
\end{lemma}
\begin{lemma}
  The input multiplicity context $\Gamma$ at index $i$ is the result of performing the monoidal operation on $x$ and the output multiplicity context $\Delta$ at index $i$.
\end{lemma}
\begin{lemma}
  For every index $j$ other than $i$, the input multiplicity context $\Gamma$ at index $i$ is the output multiplicity $\Delta$ at index $i$.
\end{lemma}
\begin{corollary}
  Given the witness $\contains{\gamma}{\Gamma}{i}{t}{x}{\Delta}$, $\Gamma \equiv \Delta$ iff $x \equiv \zero$.
\end{corollary}
\proof{All proofs are by structural induction on \textsc{VarRef}.}

\begin{nidefinition}
  The typing relation \textsc{Types} presented in Figure \ref{fig:types} is indexed over a context of types $\gamma$, a context of input multiplicities $\Gamma$, the process $P$ that is being typed, and the context of output capabilities $\Delta$, and notated as $\types{\gamma}{\Gamma}{P}{\Delta}$.
  \todo{empty output context}
  The typing rules are in one-to-one correspondence with the underlying syntax for processes, and thus we say they are syntax-directed.

  Process termination is typed under any typing context, and preserves the multiplicity context unchanged.

  Scope restriction accepts three parameters: the type $t$ of data sent over the channel; the input and output multiplicities $x$ of the data sent over the channel; and the multiplicity $y$ of the channel itself, which is used both as input and as output.
  The typing derivation of the continuation process $P$ has a channel of data type $t$ and data multiplicity $x$ added to the typing context, and the multiplicities $(y , y)$ added to the input multiplicity context.
  The continuation must completely exhaust this channel, to that end we substitute $(y , y)$ by $(\zero , zero)$ in the output context.
  \todo{comment on balancing in chan rule}

  Input requires a proof $chan_i$ locating a channel with data type $t$, data multiplicities $x$ and multiplicity $\ell_i$ \todo{} at index $i$ in the typing and multiplicity contexts $\gamma$ and $\Gamma$.
  The result of exhausting $\ell_i$ is the output multiplicity $\Xi$.
  The data type $t$ is then used to expand the typing context of the continuation $cont$, while the data multiplicities $x$ are appended to the output context $\Xi$ to form the input context over which $P$ is typed.
  These multiplicities $x$ must be exhausted by typing derivation of the continuation process.
  Note that the index $i$ is then used as part of the syntax.

  \todo{note that every new multiplicity introduced in an input context must be exhausted in the output context}

  Similar to input, output requires a proof $chan_i$ locating a channel with data type $t$ and data multiplicities $x$ in $\gamma$ at index $i$, however this time the channel must have multiplicity $\ell_o$.
  The result of exhausting $\ell_o$ is the output multiplicity $\Delta$.
  That output context $\Delta$ must have a variable of type $t$ and multiplicities $x$ at position $j$.
  This second proof represents the data that is sent over the channel.
  The result of exhausting the multiplicities $x$ of that variable is the context $\Xi$.
  The continuation process must be typed under this context $\Xi$.

  Finally, parallel composition requires its two subprocesses to be typed such that the output context of the first is the input context of the second.

  \begin{figure}[h]
  \begin{mathpar}
    \datatype{
      \gamma : \PreCtx_n \\
      \stacked{
        idxs : \Idxs_n \\\\
        \Gamma : \Ctx_{idxs}} \\
      P : \Process_n \\
      \Delta : \Ctx_{idxs}}
    {\types{\gamma}{\Gamma}{P}{\Delta} : \Set}
    \; \textsc{Types}

@@

    \inferrule
    {l : \types{\gamma}{\Gamma}{P}{\Delta} \\\\
     r : \types{\gamma}{\Delta}{Q}{\Xi}}
    {\constr{comp} \; l \; r : \types{\gamma}{\Gamma}{\comp{P}{Q}}{\Xi}}
  \end{mathpar}
  \caption{Left-over typing for a resource-aware typing system.}
  \label{fig:types}
  \end{figure}
\end{nidefinition}

@@

\todo{op but for contexts}
\begin{theorem}
  Let $\opctx{\Gamma_l}{\Delta}{\Xi_l}$ and $\opctx{\Gamma_r}{\Delta}{\Xi_r}$.
  If $\types{\gamma}{\Gamma_l}{P}{\Xi_l}$ then $\types{\gamma}{\Gamma_r}{P}{\Xi_r}$.
\end{theorem}

@@

@@

  Bidirectional typing rules needed.
